---
title: "Experimental data preparation for SwarmTCR"
author: "Maia Bennett-Boehm"
date: "`r Sys.Date()`"
output: html_document
---

# Experimental data preparation for SwarmTCR

This document describes the steps to prepare experimental data for use in SwarmTCRClassify. The experimental data should be in the form of a data frame with columns for the following variables:

- Clone ID (unique identifier for each receptor)

- CDR sequences (at least CDR1a, CDR2a, CDR3a, CDR1b, CDR2b, CDR3b)

- V and J gene segments (AV, AJ, BV, BJ)

```{r setup, include=FALSE}
source("../TCRPaired/util/library.R")
source("../TCRPaired/util/processing.R")
source("../TCRPaired/util/filtering.R")
source("../TCRPaired/util/formatting.R")
source("../TCRPaired/util/aligning.R")
source("./util/data-inputs.R")
library(rstudioapi)
knitr::opts_chunk$set(echo = TRUE)
```

The process for preparing experimental data is quite simple. The data should be placed in a single experimental data directory. This data is imported, formatted for SwarmTCRClassify (including the addition of CDR2.5 sequence if not already present), and saved as a space-delimited text file.

```{r set-variables}

experimental.path <- "./experimental-data/"
experimental.files <- list.files(experimental.path, full.names = TRUE)

```


```{r format-data}

for (file in experimental.files) {
    # Import experimental data and add CDR2.5 if not already present
    experimental.data <- import(file) %>% 
        getCDRSeq(alleles = TRUE) 
    experimental.data <- experimental.data %>% 
        formatSwarmTCR()
    file.name <- sub("\\.[[:alnum:]]+$", "", file)
    file.name <- paste0(file.name, "_CDRs.txt")
    write_delim(experimental.data, file.name)
}

```